Rectangulotopes

### 2024-04-23 draft/rectangulotopes.tex

Note that weak rectangulotopes were first constructed by S.~Law and N.~Reading in~\cite{MR2871762} as Minkowski sums of two opposite associahedra.
Our vertex and facet descriptions of \cref{thm:weakRectangulotope} can be directly derived from this construction.
Here, we provide a more pedestrian approach based on Minkowski sums of shard polytopes, as it will serve as a template for the more involved proof of \cref{thm:strongRectangulotope} in \cref{sec:strongRectangulotopes}.

@@

By construction, the quotient fan~$\c{F}_{\strongeq}$ has a maximal cone~$C_R$ for each strong rectangulation~$R$ of size~$n$.
This cone is given by~$C_R = \set{\b{x} \in \R^n}{x_i \le x_j \text{ if } i \preccurlyeq_s j}$, where~$\preccurlyeq_s$ is the strong poset of~$R$.
Note that there is a bijection between the facets of~$C_R$, the cover relations of~$\preccurlyeq_s$, and the flips from~$R$.

@@

\cref{lem:yincount} essentially gives us the coefficient $\yin{w}^R_{i,j}$ in \cref{thm:strongRectangulotope}, except that the condition $\neg i\verticalPattern j$ is replaced by $i\succ_s j$.

Recall that $i\verticalPattern j$ denote the situation where the rectangles $r_i$ and $r_j$ of inorder labels $i$ and $j$ are such that $r_i$ and $r_j$ share a vertical segment, $r_i$ is on the left of $r_j$, and the bottom edge of $r_i$ is lower than the top edge of $r_j$.

Our aim is to provide a formula for the coordinates of the vertices of the strong rectangulotope that does not involve the computation of the strong poset or one of its linear extension.
For that purpose, we prove that for every pair $i<j\in [n]$ such that $i\prec_s j$, hence which do not have any associated yin arc, we have either $cv^{T,S}_{i,j} = 0$ (hence the product in \cref{lem:yincount} is 0 as well and adding it does not harm the result), or $i\verticalPattern j$.

\begin{lemma}
  \label{lem:whentosum}
  Let $R$ be a rectangulation of size~$n$, let $S$ and $T$ be its source and target trees, and let~$([n],\preccurlyeq_s)$ be the associated strong order.
  Given a pair $i,j\in [n]$ such that $i<j$ and $i\prec_s j$, we have
  \[
    cv^{T,S}_{i,j} > 0 \quad \Longrightarrow \quad i \verticalPattern j.
  \]
\end{lemma}
\begin{proof}
  Consider the two rectangles $r_i$ and $r_j$ with inorder labels $i$ and $j$ in $R$.
  We first observe that the source vertex $s(r_j)$ is not contained in the interior of the lower left quadrant of origin $t(r_i)$.
  Indeed, if we suppose otherwise, then the line segment from $s(r_j)$ to $t(r_i)$ has positive slope, which implies that $j\tri i$, which in turn implies $j\prec_s i$, contradicting our hypothesis.

  We now claim that there can only be one segment on which leaves common to the vertical subtree of $i$ in $T$ and the vertical subtree of $j$ in $S$ may lie, and this segment is the vertical segment supporting the right edge of $r_i$ and the left edge of $r_j$.
  \Jean{This requires a case analysis. Any shortcut?}

  Let us assume that this segment is shared.
  If the bottom edge of $r_i$ is higher than the top edge of $r_j$, we must have $j\btri i$, which implies $j\prec_s i$, contradicting our hypothesis again.
  Hence we must have that the bottom edge of $r_i$ is lower than the top edge of $r_j$, hence that $i\verticalPattern j$, as claimed.
\end{proof}

We omit the proofs of the analogues of \cref{lem:yincount} and \cref{lem:whentosum} for the case of the yang arcs: The arguments are completely symmetric.
This concludes the proof of \cref{thm:strongRectangulotope}.

>>>>

Note that weak rectangulotopes were first constructed by S.~Law and N.~Reading in~\cite{MR2871762} as Minkowski sums of two opposite associahedra.
Our vertex and facet descriptions of \cref{thm:weakRectangulotope} can be directly derived from this construction.
Here, we provide a more pedestrian approach based on Minkowski sums of shard polytopes, as it will serve as a template for the more involved proof of \cref{thm:strongRectangulotope} in \cref{sec:strongRectangulotopes}.
\jean{From \cref{thm:weakCong}, we obtain a bijection between the cover relations in the weak poset and the facets of the dual cone corresponding to the rectangulation, which are in turn one-to-one with feasible flips in the rectangulation.}

@@

By construction, the quotient fan~$\c{F}_{\strongeq}$ has a maximal cone~$C_R$ for each strong rectangulation~$R$ of size~$n$.
This cone is given by~$C_R = \set{\b{x} \in \R^n}{x_i \le x_j \text{ if } i \preccurlyeq_s j}$, where~$\preccurlyeq_s$ is the strong poset of~$R$.
Note that there is a bijection between the facets of~$C_R$, the cover relations of~$\preccurlyeq_s$, and the flips from~$R$.
\jean{Note that, similarly to the weak case, \cref{thm:strongCong} yields a bijection between the cover relations in the strong poset and the facets of the dual cone corresponding to the rectangulation, which are in turn one-to-one with feasible flips in the rectangulation.}

@@

\cref{lem:yincount} essentially gives us the coefficient $\yin{w}^R_{i,j}$ in \cref{thm:strongRectangulotope}, except that the condition $\neg i\verticalPattern j$ is replaced by $i\succ_s j$.
Recall that $i\verticalPattern j$ denote the situation where the rectangles $r_i$ and $r_j$ of inorder labels $i$ and $j$ share a vertical segment, $r_i$ is on the left of $r_j$, and the bottom edge of $r_i$ is lower than the top edge of $r_j$.
Our aim is to provide a formula for the coordinates of the vertices of the strong rectangulotope that does not involve the computation of the strong poset or one of its linear extension.
For that purpose, we prove that for every pair $i<j\in [n]$ such that $i\prec_s j$, which do not have any associated yin arc, we have either $cv^{T,S}_{i,j} = 0$ (hence the product in \cref{lem:yincount} is 0 as well and adding it does not harm the result), or $i\verticalPattern j$.

\begin{lemma}
  \label{lem:whentosum}
  Let $R$ be a rectangulation of size~$n$, let $S$ and $T$ be its source and target trees, and let~$([n],\preccurlyeq_s)$ be the associated strong order.
  Given a pair $i,j\in [n]$ such that $i<j$ and $i\prec_s j$, we have
  \[
    cv^{T,S}_{i,j} > 0 \quad \Longrightarrow \quad i \verticalPattern j.
  \]
\end{lemma}
\begin{proof}
  Consider the two rectangles $r_i$ and $r_j$ with inorder labels $i$ and $j$ in $R$.
  We first observe that the source vertex $s(r_j)$ is not contained in the interior of the lower left quadrant of origin $t(r_i)$.
  Indeed, if we suppose otherwise, then the line segment from $s(r_j)$ to $t(r_i)$ has positive slope, which implies that $j\tri i$, which in turn implies $j\prec_s i$, contradicting our hypothesis.
  We also observe that since $i < j$, by definition of the inorder, the rectangle $r_j$ must be below or on the right of $r_i$.
  Combining those two observations, we conclude that the left edge of $r_j$ cannot be on the left of the right edge of $r_i$.

  Therefore, there can only be one segment on which a leaf common to the vertical subtree of $i$ in $T$ and the vertical subtree of $j$ in $S$ may lie, which is the vertical segment supporting the right edge of $r_i$ and the left edge of $r_j$.
  Let us assume that this segment is shared by those two edges.
  If the bottom edge of $r_i$ is higher than the top edge of $r_j$, we must have $j\btri i$, which implies $j\prec_s i$, contradicting our hypothesis again.
  Thus the bottom edge of $r_i$ is lower than the top edge of $r_j$, and $i\verticalPattern j$, as claimed.
\end{proof}

We omit the proofs of the analogues of \cref{lem:yincount} and \cref{lem:whentosum} for the case of the yang arcs, which use completely symmetric arguments.
This concludes the proof of \cref{thm:strongRectangulotope}.
